---
title: "BAN421-report"
author: "Group 2"
date: "November 22, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
1. Introduction

In this paper, we try to investigate the influence of other countries' total supply (production)
and total consumption on the electricity market spot prices in Norway, investigating prices for Bergen and Tromsø. All the input data is taken from the Nordpool webiste and is dating from 2014 to 2018. 

# Economic reasoning 
Comparing production and consumption volumes of electricity for a country, one can easily calculate the delta of those, representing times in which there is either overproduction or oversupply. Assuming a non-perfect energy storage capacitiy, these peaks and lows in the nation-wide electricty net provide the opportunity and need to trade electricity with economically and geographically closely related countries [citation].

Therefore, our Reasearch Question is: how depended is Norway and its electricity price from other nordic countries and their electricity production?

Before starting the analysis, we explore the most efficient way to accommodate multiple files when loading raw data into an R environment, specifically, by comparing the processing time using dataframe and tibble. After that, we will make plots to see the general consumption and production trends for each country during the past five years. In addition, the price trend in Norway will also be plotted to see how price had fluctuated during the past five years. Under the help of these plots, we will have a better understanding about the evolution of each factor. In order to investigate the relationship between consumtion, production and price, we will use baysian statistics, and specifically the Hidden Markov Model (HMM). We will implement it with a suitable R package and finally interpret the results.

# If we are really going ahead with the Dplyr package (since this guy HATES it, we should at least acknowledge that this is not the only way)
For the purpose of our investigation, we decided to use packages from the so-call R "tidyverse". While the group is aware of the multiple downsides of using piped statements, we decided for the purpose of this particular investigation to make use of them.

```{r Libraries, warning = FALSE, message = FALSE, echo = FALSE}
# loading required packages
library(readxl)
library(tibble)
library(quantmod)
library(MASS)
library(reshape2)
library(dplyr) ## really?
library(ggplot2)
library(gridExtra)
library(magrittr) ## really?
library(depmixS4)
# rm( list=ls() )
```
# Preprocessing and analysis
2. Preparing for analysis

For our first step, we use system.time to compare the processing time of using dataframe and tibble to accomadate the data we cleaned. We get the following result:
```{r Data cleaning, warning = FALSE, message = FALSE, echo = FALSE}
## 1.Data cleaning process ----

#find the file names of consumption, spot prices and production for each year
consum_names <- list.files(path = ".", pattern = "consumption-per.*.xlsx", full.names = T)
elspot_names <- list.files(path = ".", pattern = "elspot.*.xlsx", full.names = T)
production_names <- list.files(path = ".", pattern = "production-per.*.xlsx", full.names = T)

# try to load all 5 years consumption data into one dataframe
consumption <- data.frame() # create one empty dataframe first
system.time(for(i in consum_names){
  #since our data is  pure xls format, read_xls fits the most
 consumption <- read_xlsx(i)%>% 
       bind_rows(consumption)%>% #combine each year's data together by rows
   na.omit()
})
```
Compare against t
```{r Time comparison, warning = FALSE, message = FALSE, echo = FALSE}
#use tibble to compare the processing time
consumption_tib<-tibble()
system.time(for(i in 1:length(consum_names)){
  consumption_tib <- read_xlsx(consum_names[i])%>% 
    bind_rows(consumption_tib)  
})
```
We cannot observe a significant difference between the processing times of either using tibble or dataframe. This might be due to the fact that our raw data is not large enough to observe a difference between the different loading algorithms.
 
To further compare the difference between dataframe and tibble, we also look at the structure of results produced by each. We get the following result:

```{r Check data structure, warning = FALSE, message = FALSE, echo = FALSE}
#check the data structure in each
str(consumption) 
```
```{r, warning = FALSE, message = FALSE, echo = FALSE}
str(consumption_tib)
```

They have the same data structure, so we will continue to use dataframe to load production and price data.

```{r Binding dataframes, warning = FALSE, message = FALSE, echo = FALSE}
production <-data.frame()
for(i in 1:length(production_names)){
  production <- read_xlsx(production_names[i])%>% 
    bind_rows( production)%>%
    na.omit()
}

price <-data.frame()
for(i in 1:length(elspot_names)){
  price<- read_xlsx(elspot_names[i])%>% 
    bind_rows( price)
}
price<-price[,-5]
price<-na.omit(price)
```

We continue to load the prognosis data for consumption and production in the different countries, so that we can use it to predict based on our model.

```{r Finalized R-data, warning = FALSE, message = FALSE, echo = FALSE}
consum_pre<-list.files(path = ".", 
                           pattern = "consumption-pro.*.xlsx", full.names = T)
consum_pre<-read_xlsx(consum_pre)
produc_pre<-list.files(path = ".", 
                           pattern = "production-pro.*.xlsx", full.names = T)
produc_pre<-read_xlsx(produc_pre)
# sum up all productions from small areas in each country and take the total only
produc_pre<- produc_pre%>%
  mutate(NO=NO1+NO2+NO3+NO4+NO5)%>%
  mutate(SE=SE1+SE2+SE3+SE4)%>%
  mutate(DK=DK1+DK2) %>%
  dplyr::select("Date","Hours","FI","DK","EE", # specify to avoid overlap with "MASS" package
         "LV","LT","NO","SE")

# save all these data as R.data
save(consum_pre,consumption,price,produc_pre,production,file="BAN421.Rdata")
```

3. Plotting

The next step in our research is to create plots in order to have a general picture about price,production, and consumption changes over the five year timeframe. After creating the yearly amounts, we firstly look at the variety in consumption in Norway.

```{r Data Plotting, warning = FALSE, message = FALSE, echo = FALSE}
#load("BAN421.Rdata") # load data, if start only from here

# create another "year" variable 
consumption<-consumption%>%
  mutate(year = format(Date, "%Y"))
production<-production%>%
  mutate(year = format(Date, "%Y"))
price<-price%>%
  mutate(year = format(Date, "%Y"))

# i) consumption
# create the yearly plotting data
plot_c<-consumption %>%
  group_by(year) %>%
  dplyr::select("year","NO","SE","DK","FI","LV","LT")%>%
  summarise_each(funs(sum))

# plot consumtion trend in Norway 
ggplot(data=consumption,mapping=aes(x=consumption$Date))+
  geom_line(aes(y=consumption$NO, color="NO"))+
  xlab("Date")+
  ylab("consumption")+
  ggtitle("Daily Consumption in Norway")
```

One can observe seasonal fluctuations, with a higher demand for electricity in winter months than in summer months. Following, we compare the yearly consumption of all countries.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
#and all countires
ggplot(data=plot_c,mapping=aes(x=year))+
  geom_point(aes(y=plot_c$NO, color="NO"))+
  geom_point(aes(y=plot_c$SE,color="SE"))+
  geom_point(aes(y=plot_c$DK,color="DK"))+
  geom_point(aes(y=plot_c$FI,color="FI"))+
  geom_point(aes(y=plot_c$LV,color="LV"))+
  geom_point(aes(y=plot_c$LT,color="LT"))+
  ggtitle("Yearly consumption for all countries")+
  ylab("Total electricity consumption")
## can we change to smth. that shows more "meaning"/ with a line?
```

Here one can see that Norway and Sweden have the highest yearly consumption. The trend for all countries is slightly increasing over the years, but mostly constant. Following, we plot the same for the production.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# 2. Production
# create the yearly plotting data
plot_p<-production %>%
  group_by(year) %>%
  dplyr::select("year","NO","SE","DK","FI","LV","LT")%>%
  summarise_each(funs(sum)) #sum the toal production each country each year
# use ggplot to plot production trend in Norway 
ggplot(data=production,mapping=aes(x=production$Date))+
  geom_line(aes(y=production$NO, color="NO"))+
  xlab("Date")+
  ylab("production")+
  ggtitle("Daily Production in Norway")
```
```{r, warning = FALSE, message = FALSE, echo = FALSE}
# ...and for all  countries
ggplot(data=plot_p,mapping=aes(x=year))+
  geom_point(aes(y=plot_p$NO, color="NO"))+
  geom_point(aes(y=plot_p$SE,color="SE"))+
  geom_point(aes(y=plot_p$DK,color="DK"))+
  geom_point(aes(y=plot_p$FI,color="FI"))+
  geom_point(aes(y=plot_p$LV,color="LV"))+
  geom_point(aes(y=plot_p$LT,color="LT"))+
  ggtitle("Yearly production for all")+
  ylab("production")
## can we change to smth. that shows more "meaning"/ with a line?
```

Also here we can observe just slighty fluctuating volumes, with Norway and Sweden being the main producers. Now, we have a look at the development of the electricity price in Norway for two cities.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# 3.Price

# Use ggplot to plot price trend in Norway 
ggplot(data=price,mapping=aes(x=price$Date))+
  geom_line(aes(y=price$Tromsø, color="Tromsø"))+
  geom_line(aes(y=price$Bergen, color="Bergen"))+
  xlab("date")+
  ylab("daily price")+ #what type of price? spot, average..?
  ggtitle("Daily Price in Norway")
```

Quite interesting to observe, there are heavy fluctuations in the electricity price, and for a few times, there are heavy spikes. Furthermore, prices in Bergen and Tromsø do differ.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# Create the yearly plotting data
plot_price<-price %>%
  group_by(year) %>%
  dplyr::select("year","Bergen","Tromsø")%>%
  summarise_each(funs(mean))#calculate the mean price in Norway each year in each place
# A general trend in each year 
ggplot(data=plot_price,mapping=aes(x=year))+
  geom_point(aes(y=plot_price$Bergen, color="Bergen"))+
  geom_point(aes(y=plot_price$Tromsø, color="Tromsø"))+
  ggtitle("Yearly price in Norway")+
  ylab("mean_price")
## can we change to smth. that shows more "meaning"/ with a line?
```
From the above figures, we can see that the consumption and production in Norway differentiate from each other quite a lot, which means that Norway has been trading (import and export) electricity with other countries. This gives us a positive signal about our hypothesis that other countries' consumption and production may have an impact on the price in Norwegian market. The market spot price in Norway has been fluctuating since 2014 and presents an incresing trend since 2015. Especially, there are two peak time periods in Tromsø. The reasons behind this are not that clear and we belive that there are way more factors influencing this. However in our report, we are going to focus only on the impact by other countries' consumption and production.

3. Model: Introduction, definition, fit

# Theory
Therefore, in the following we will introduce the Hidden Markov Model (HMM) to uncover this. The HMM can be seen as the [... + sources]
In economics, HMM are mainly applied as so-called regime switching models.

# Packages
To indentify suitable packages to implement this method in R, we looked through the "Comprehensive R Archive Network Task View" and found in total three packages: "depmixS4", "HMM", and "HiddenMarkov". After comparing and testing, we decided for "depmixS4", which is regulary updated and often used in recent research (e.g. CITATION). [...]

# Approach 
To prepare data for the HMM model, we need to define the different states we are interested in, which are the other countries and which influence they have on the Norwegian electricity price. Therefore, we take the delta of supply and demand for a given country, where a positive number indicates an over-production and therefore an export to another country (e.g. Norway) and a negative number indegates an under-consumsion, which means that the country need to import electricity. Therefore, we define:
S1: 
S2: 

```{r Statistical methods, message=FALSE, warning=FALSE, include=FALSE}
## 3.Statistical methods ----
# Combine tables with all the data and preproceess the data----
comb <-  inner_join(consumption, production,
                   by = c("Date", "Hours"))

# Add new parameters to the combined table based on the production and consumption parameters
comb <- comb %>% 
  mutate(NO = NO.y - NO.x,
         DK = DK.y - DK.x,
         SE = SE.y - SE.x,
         NORD = Nordic.y - Nordic.x)

# Clean the table
comb <- comb[c("Date", "Hours", "NO", "DK", "SE", "NORD")]

# Summarise by date to cancel minor chages and noise
comb <- comb %>% 
  group_by(Date) %>% 
  summarise(NO = sum(NO),
            DK = sum(DK),
            SE = sum(SE),
            NORD = sum(NORD))

# Add logical parameters showing the state of each country
comb <- comb %>% 
    mutate(NO_state = as.logical(NO > 0),
           DK_state = as.logical(DK >0),
           SE_state = as.logical(SE > 0),
           NORD_state = as.logical(NORD > 0))

comb <- as.data.frame(lapply(comb, function(y) gsub("TRUE", "Surplus", y)))
comb <- as.data.frame(lapply(comb, function(y) gsub("FALSE", "Deficit", y)))

# Convert to the proper formatting
comb[6:9] <- lapply(comb[6:9], as.character)
comb$Date <- as.Date(comb$Date)

# Run and fit HMM----
hmm <- function (param, data){
  
  # We will use the changes in fluctuations as dependece ground for the HMM
  fluc <- diff(log(as.numeric(param)))
  fluc <- c(0, fluc)
  
  # Set the model
  model <- depmix(fluc ~ 1, 
                family = gaussian(), # normal (gaussian) param distribution
                nstates = 2, # number of states
                data = comb) # used data 
  
  # Fit the data and get it as data frame with the defined states
  fit.hmm <- fit(model)
  est.states <- posterior(fit.hmm)
  
  # Analyse the results
  tbl <- table(est.states$state, comb$NO_state)
  colnames(est.states)[2:3] <- c("Deficit", "Surplus")
  est.states$Date <- comb$Date
  
  # Change the naming of the states based on the tbl
  grep("Deficit", comb$NO_state)
  grep("1", est.states$state)
  
  est.states$state <- gsub("1", "Deficit", est.states$state)
  est.states$state <- gsub("2", "Surplus", est.states$state)
  
  return(est.states)
}

# Plot the results----
display.hmm <- function(original, 
                        model){
  
  # Define colours
  mycols <- c("darkmagenta", "turquoise")
  
  # Plotting the actual states for the further comparison with the HMM estimates (we look only into Norwegian states)
  g1 <- ggplot(original, aes(x = Date, y = NO_state, fill = NO_state, col = NO_state)) + 
    geom_bar(stat = "identity", alpha = I(0.75)) + 
    scale_fill_manual(values = mycols, name = "State:\nProduction of\nNorway", labels = c("Deficit", "Surplus")) +
    scale_color_manual(values = mycols, name = "State:\nProduction of\nNorway", labels = c("Deficit", "Surplus")) +
    theme(axis.ticks = element_blank(), axis.text.y = element_blank()) +
    labs(y = "Actual State")
  
  # Visualise the HMM model estimates
  g2 <- ggplot(model, aes(x = Date, y = state, fill = state, col = state)) + 
    geom_bar(stat = "identity", alpha = I(0.75)) +
    scale_fill_manual(values = mycols, name = "State:\nProduction of\nNorway", labels = c("Deficit", "Surplus")) +
    scale_color_manual(values = mycols, name = "State:\nProduction of\nNorway", labels = c("Deficit", "Surplus")) +
    theme(axis.ticks = element_blank(), axis.text.y = element_blank()) + 
    labs(y = "Estimated State")
  
  return(grid.arrange(g1, g2,
                      widths = 1, 
                      nrow = 2))
}
```

4. Interpretation 

The final graph shows us [...]